Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-02-20 paper1_jiturra.Rmd
parrafo class redes

Apart from the individual approach to attitude formation, we argue that people also form their opinions in the economic domain on the basis of their social relations. Despite the limited research on the link between social ties and attitudes in the economic domain, we identify two broad theoretical approaches that have discussed the role of social networks in attitude formation.

On the one hand, one approach accentuates the role of beliefs about economic inequality rooted in social comparison processes as an explanation through which people form their opinions about redistributive policies [@condon_economic_2020]. This hypothesis can be traced to the studies on class images and perceived class conflicts [@Evans1992;@KelleyEvans1995]. Here, the initial argument is that people form their beliefs through the individual, family, friends, and coworkers experiences instead of the whole society, which is described as an availability heuristic that systematically biases inferences about inequality based on the homophily of reference groups [@Evans1992, p. 467]. From this perspective, how people infer the social world is linked to the degree of segregation in their immediate social environment, which influences the intensity and character of the information, ultimately shaping inequality perceptions [@mijs_is_2021]. Thus, experience sharing in conversations with socioeconomically diverse networks has been proved to contribute to the accuracy of the images of income and wealth inequalities in contrast to people in more isolated discussion networks [@summers_deliberating_2022]. Despite this, we argue that this research field has been more focused on the cognitive dimension of preference formation through inequality perceptions, either relying on surveys [e.g. @cansunar_who_2021;@garcia-castro_perceived_2022] or experimental manipulations [e.g. @Cruces2011;@becker_significant_2021], rather than empirically addressing the role of class segregation in social networks and its claimed influence on attitudes in the economic domain.

On the other hand, a different approach has suggested that the observed patterns of social ties through a network approach provide a broader image of class relations that ultimately contribute to the consolidation of social norms and group identity [@kalmijn_social_2007, p. 550]. This claim resembles the fact that classes are characterized as collectivities with differences in their degrees of cohesion and solidarity, encompassing unequal status-based social interactions that are linked to the individual or household material well-being, cultural perspectives, and political preferences that structure the broader social experiences [@morris_attenuation_1996, p. 48].

More direct evidence on social ties, class, and redistributive preferences in general populations will be discussed here. First, starting with the case of the United States [@lee_consider_2023], then Sweden [@lindh_missing_2021], and cross-country comparisons in Europe [@paskov_crossclass_2022]. Additionally, studies on network segregation on preferences for market distribution of social services [@beck_explorando_2019] and attitudes towards inequality in general [@otero_power_2023].

>>>>

Apart from the individual approach to attitude formation, we argue that people also form their opinions in the economic domain based on their social relations. Despite the limited research on the link between social ties and attitudes in the economic domain, we identify two broad theoretical approaches that have discussed the role of the social environment in attitude formation.

On the one hand, one approach accentuates the role of beliefs about economic inequality rooted in social comparison processes as an explanation through which people form their opinions about redistributive policies [@condon_economic_2020]. This hypothesis can be traced to the studies on class images and perceived class conflicts [@Evans1992;@KelleyEvans1995]. Here, the initial argument is that people form their beliefs through the individual, family, friends, and coworkers experiences instead of the whole society, which is described as an availability heuristic that systematically biases inferences about inequality based on the homophily of reference groups [@Evans1992, p. 467]. From this perspective, how people infer the social world is linked to the degree of segregation in their immediate social environment, which influences the intensity and character of the information, ultimately shaping inequality perceptions [@mijs_is_2021]. Thus, experience sharing in conversations with socioeconomically diverse networks has been proven to contribute to the accuracy of the images of income and wealth inequalities compared to people in more isolated discussion networks [@summers_deliberating_2022]. Despite this, we argue that this research field has been more focused on the cognitive dimension of preference formation through inequality perceptions, either relying on surveys [e.g. @cansunar_who_2021;@garcia-castro_perceived_2022] or experimental manipulations [e.g. @Cruces2011;@becker_significant_2021], rather than empirically addressing the role of class segregation in social networks and its claimed influence on attitudes in the economic domain.

Differently, it has been suggested that the lens of a network approach provides a better picture of the class relations that nurture social norms and group identity [@kalmijn_social_2007, p. 550]. This claim resembles the fact that classes are characterized as collectivities with differences in their degrees of cohesion and solidarity, encompassing unequal status-based social interactions that are linked to the individual or household material well-being, cultural perspectives, and political preferences that structure the broader social experiences [@morris_attenuation_1996, p. 48]. Furthermore, these social relations contribute to the degree to which people reflect on the collective, so that in societies where opportunities for contact between people of different social classes can be detrimental to social integration. It has been argued that the social distance among groups widened by extreme inequality can be described as an _empathy gulf_ that comprises barriers to imagining others’ lifestyles and goods, which can also be expressed in spatially segregated interactions and isolated lifeworlds that provide meaning to social inequality experiences, resulting in doubts about worthiness when the lower classes contrast themselves with extravagant lifestyles from the upper classes, undermining feelings of social inclusion and cohesion [@sachweh_moral_2012]. Consequently, segregation drives the lives of others to become more distant and might have consequences on empathy and solidarity toward others, potentially leading to perceiving fellow citizens as strangers [@otero_lives_2022, p. 758]. Thus, we suggest that attitudes towards economic inequality may be influenced by the class situation of both the individual and network ties, resembling the core claims of intergroup contact theory [@pettigrew_metaanalytic_2006] so that social influence processes can either align or polarize attitudes according to the characteristic of groups as well as their contact opportunities given the degree of segregation.

In recent years, we have seen an increase in this _network turn_ of class analysis, contributing to unraveling the relationship between the composition of social ties and attitudes toward inequality. In this regard, it has been shown that the degree of contact between different classes has shown that individuals of the lower class interact with those of upper classes may consider economic inequality as justifiable, whereas interactions with those of the lower class may prompt individuals from different social classes to question the fairness of income distribution [@vargassalfate_contact_2023]. Further evidence suggests that class-based contact diversity shortens the social distance between social classes, providing broader images about the living conditions of others, which can trigger greater concerns about inequality [@otero_power_2023] and undermine support for the market distribution of social services [@beck_explorando_2019].

In addition, studies that have more directly addressed the relationship between networks and social class have been consistent with previous findings. In this regard, evidence from Sweden suggests that people tend to assimilate their opinions according to their surrounding friends and acquaintances, which is reinforced by class homophily, where higher contacts in the managerial class negatively impact redistributive demands, while ties to the working class increase them arguing that “individuals take an impression from others and modify their attitudes accordingly” [@lindh_missing_2021, p. 698]. Similarly, additional socialization sources are cross-class ties through the family members. Family of origin is crucial for political socialization, where class interests and norms are nurtured in childhood and early adulthood. A study in the United States has shown that those tied to the upper class through parental relations are less supportive of redistribution and progressive taxation than those with working-class family backgrounds [@lee_consider_2023]. Besides, @paskov_crossclass_2022 argues that the class position of family and partners shapes preferences en tanto que los households share risk according to the class position of its members, where class homogeneity in ties polarizes preferences between lower and upper classes. In contrast, this gradient is blurred as heterogeneity increases according to the class position of the partner and family of origin.

In sum, the attitudes of individuals toward redistribution are influenced by their class-embedded social relations. Thus, the degree of class segregation in these networks plays a crucial role in shaping people's beliefs about income distribution and redistributive policies. Furthermore, the degree of contact between different classes and class homophily reinforces the assimilation of opinions according to surrounding friends and acquaintances. Therefore, we expect that being segregated in similar class environments polarizes own class interests, where segregated lower (upper) classes hold more (less) redistributive preferences.
